Effect of resonant and random excitation on the LHC beam in the context of pulsed hollow electron lens operation.

### 2018-02-15 resex.tex
Changes from Giulio completed

\section{Sources of residual field in the proton beam core region and first order estimates\label{sec:core}}
With the current HEL layout as shown in Fig.~\ref{fig:hel_layout}, parasitic kicks on the proton beam core can arise in the central region due to profile imperfections in the electron beam and at the entrance and exit of the HEL, where electron and proton beam overlap. Currently, no HEL is installed in the LHC, and the kick on the proton beam core must therefore be emulated with the help of other devices already installed in the LHC. For this purpose the kick is approximated to first order by a dipole kick. During the experiments presented in this paper, the dipole kick with the corresponding excitation pattern could then be applied with the LHC transverse damper as will be described in detail in Section~\ref{sec:adt}. As will be shown in Sec.~\ref{core:sec:1} and \ref{core:sec:2}, the expected kick to first order from the HEL bends is estimated to be~(Eq.~\ref{eqn:kick_bends}):
\begin{equation}
	\Delta x'_{\rm{bends}},\Delta y'_{\rm{bends}}\approx \SI{0.5}{nrad},
\end{equation}
and for the central region due to profile imperfections~(Eq.~\ref{eqn:kick_central}):
\begin{equation}
	\Delta x'_{\rm{central \ region}}, \Delta y'_{\rm{central \ region}} \approx \SI{19.6}{nrad} % this should be rounded to the same number of digits as the 0.5nrad
\end{equation}
where the current HEL design parameters are used (see Table~\ref{tab:hllhc_param}--\ref{tab:hel_param}). The contribution from the central region is clearly dominating.

@@

Using these measurements and calculations of the relative electric field strengths in the transverse plane (Fig.~\ref{core:fig:1}), we obtain an average field in the hole of
\begin{equation}
	E_{\mathrm{rel}}= 5\%
\end{equation}
For a maximum kick of $392$~nrad as derived in Eq.~\ref{eqn:helkick}, the estimated dipole kick amplitude at the proton beam core then is:
\begin{equation}\label{eqn:kick_central}
\Delta x'_{\rm{central \ region}}, \Delta y'_{\rm{central \ region}} = \SI{19.6}{nrad}
\end{equation}

@@

The ADT kicker deflection angle is calculated from the kicker geometry and the excitation voltage on the deflection plates. The excitation voltage used in the experiments is a complicated waveform, generated by a very complex hardware (digital signal processor, chain of transmission lines, low and high power amplifiers). The peak excitation voltage (and subsequently the kick) seen by the beam can be only indirectly measured using the probes on each kicker. The measured values are therefore believed to be known with an error margin of 10-15\% for the experiments in 2017 and 50\% in 2016. The estimate for 2017 is based on only a single measurement and in the future and with more time available the determination of the error would definitely have to be improved. The error in 2016 is estimated from the transverse damper operational parameters. Due to lack of the dedicated machine time no precision calibration was performed. The maximum kick strength reachable with the ADT at injection without risking saturation is 96~nrad. As an example Fig.~\ref{fig:fill_meas} shows the oscillation amplitude of each bunch in both H and V planes captured by the LHC real time transverse activity monitor during the 2017 experiments and an excitation with $8^{\mathrm{th}}$ turn pulsing. The measured activity is well in line with the expected excitation pattern.
\begin{figure}[h]
	\centering
	\includegraphics[width=1.0\linewidth]{bunchfilling_measured.png}	
	\caption{\label{fig:fill_meas} Bunch by bunch excitation detected by the real time transverse activity monitor during the 2017 experiment and an excitation with 8th turn pulsing.}
\end{figure}

@@

The sensitivity to $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing is also observed in absence of machine errors in which case the only source of non-linearities are sextupoles and octupoles~\cite{md_sim_hel_res_ex_fitterer} suggesting that latter are responsible for the observed sensitivity. The driven resonances are revealed by the FMA analysis shown in Fig.~\ref{fig:patternfma}. The $7Q_x$ resonance is excited in case of the $7^{\mathrm{th}}$~turn pulsing and the $10Q_x$ and $10Q_y$ resonance in case of the $10^{\mathrm{th}}$~turn pulsing. As octupoles can only drive even resonances, the source of the $7Q_x$ resonances are the sextupoles while the octupoles only generate the tune footprint. The other pulsing patterns do not exhibit any increase in losses or emittance growth without magnetic errors and their effect can be thus attributed to the magnetic field errors, implying also the seed chosen for the simulation.
\begin{figure}[h]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		no excitation
		\includegraphics[width=1.0\linewidth]{2016injnocolc15o+19_6noerru_dp0_ord10_annotate.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$10^{\mathrm{th}}$ H+V
		\includegraphics[width=1.0\linewidth]{2016injnocolc15o+19_6noerrut10skhv_dp0_ord10.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$7^{\mathrm{th}}$ H
		\includegraphics[width=1.0\linewidth]{2016injnocolc15o+19_6noerrut7skh_dp0_ord7.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$7^{\mathrm{th}}$ V
		\includegraphics[width=1.0\linewidth]{2016injnocolc15o+19_6noerrut7skv_dp0_ord7.png}
	\end{minipage}
	\caption{\label{fig:patternfma} FMA analysis without excitation (top left) for $10^{\mathrm{th}}$~turn pulsing in H+V (top left) and for $7^{\mathrm{th}}$~pulsing only in H (bottom left) and only in V (bottom right) and  based on the 2016 injection optics with no machine errors and a tune of (64.28,59.31). The excitation is 120~nrad in the corresponding plane. The absence of a strong excitation of any resonance in case of $7^{\mathrm{th}}$~turn pulsing only in V and the strong excitation in case of pulsing only in H confirms the excitation of the $7Q_x$ resonance. For $10^{\mathrm{th}}$~turn pulsing there is in contrast no significant difference between pulsing only in H, only in V or in H+V (see Appendix~\ref{app:sec:10}, Fig.~\ref{app:fig:fma:10}).}
\end{figure}

One may have noticed that the emittance in case of the $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing starts at an increased initial value and then stays almost constant during the entire simulation time, which usually points to an error in the chosen simulation setup. This behavior is however typical for any resonant excitation and can be ascribed to the adjustment of the beam distribution during the first $10^4$ turns to a new equilibrium. As the beam distribution is only dumped every $10^4$ turns, this initial fast adjustment manifests itself in an increased initial emittance value in Fig.~\ref{fig:patternsim}, and also any other simulations over $10^6$ turns presented in this paper. Besides an increased emittance, the new distribution also becomes non-Gaussian. This is illustrated by means of the residual in respect to the Gaussian distribution for $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing in Fig.~\ref{fig:patternhist}. This initial fast change of the beam distribution is not an artifact of the simulation, but has been, to the knowledge of the authors, also for the first time experimentally observed during the 2016 and 2017 LHC experiments presented in this paper. The timescale of this fast adjustment is in experiments however much slower than predicted in simulations --- on the timescale of minutes compared to seconds. 
\begin{figure}[t]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$7^{\mathrm{th}}$ H+V
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_t7skhv_3_5um_hist_x.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$10^{\mathrm{th}}$ H+V
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_t10skhv_3_5um_hist_y.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		random H+V
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_ranhv_3_5um_hist_x.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		no excitation
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_3_5um_hist_y.png}
	\end{minipage}	
	\caption{\label{fig:patternhist} Horizontal beam distribution for $7^{\mathrm{th}}$~turn pulsing (top left), vertical beam distribution for $10^{\mathrm{th}}$~turn pulsing (top right) and horizontal (bottom left) and vertical (bottom right) beam distribution for random excitation from distribution tracking simulations based on the 2016 injection optics with $(Q_x,Q_y)=(64.28,59.31)$ and standard errors. The excitation is applied in both planes with an amplitude of 96~nrad. The residual is defined as the final distribution (here after $10^4$~turns) minus the initial distribution (here after $10^2$~turns) as \%, and the ratio as the final distribution divided by the initial distribution. The red line in the plot of the residual is the difference between the Gaussian fit of the distribution and the distribution itself.}
\end{figure}

@@

\subsection{$10^{\mathrm{th}}$ turn pulsing\label{sec:simex10}}
The $10^{\mathrm{th}}$~turn pulsing pattern was tested in 2016 with an excitation in the vertical plane only. As this was the first time a resonant excitation was tested in the LHC, only in total 48~bunches could be injected in order to still guarantee the protection of the machine. To test different excitation amplitudes during one fill and have enough statistics, the excitation was therefore only applied in one plane. As described earlier, the experiment started with a period of 12~minutes without excitation in order to let the beam distribution fully adjust to its equilibrium state after injection. Afterwards the excitation was applied with the excitation scheme shown in Fig.~\ref{fig:fill} and a maximum excitation amplitude of $A_{\mathrm{max}}=5\Delta A = 48~$nrad for a total of 11~minutes. The excitation was then further increased to the maximum value reachable without saturation of $A_{\mathrm{max}}=5\Delta A = 96~$nrad and kept for 11~minutes. In both cases, the excitation induced the following changes:
\begin{enumerate}
	\item Increase of the loss rate with the excitation amplitude (Figs.~\ref{fig:10thexp}.
	\item Increase of the emittance growth with the excitation amplitude in the vertical plane but not in the horizontal plane (Fig.~\ref{fig:10thexp}).
	\item Change of the beam distribution for the excited bunches (Fig.~\ref{fig:10thexpprof}).	
\end{enumerate}
As loss rates, emittance growth and beam distribution stayed unchanged for the $4$~reference bunches, the above observations can be also truly associated with the effect of the $10^{\mathrm{th}}$~turn pulsing.
\begin{figure*}[h]
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[height=0.75\linewidth]{2016_bunch_intensity_v10th_no_damper_avg.png}
	\end{minipage}
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[height=0.75\linewidth]{2016_emith_avg_rel_v10th_no_damper.png}
	\end{minipage}	
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[height=0.75\linewidth]{2016_emitv_avg_rel_v10th_no_damper.png}
	\end{minipage}	
	\caption{\label{fig:10thexp} Relative intensity (top) and relative horizontal (bottom left) and vertical (bottom right) emittance averaged over the bunches experiencing the same excitation amplitude during the 2016 experiments. For all bunches the damper was not active. Relative refers here to the normalization to the initial value. For each maximum amplitude $A_{\mathrm{max}}=48$~nrad and 96~nrad indicated with black arrows, the excitation amplitude was linearly increased for each group of 4~bunches (see Fig.~\ref{fig:fill}). The different colored solid lines labeled with $n\Delta A$ are the average over the 4~bunches with the same excitation amplitude together with the $1\sigma$ standard deviation shown as envelope. The area with a blue background indicates the fast adjustment period of the beam distribution transitioning into a new equilibrium state highlighted with a gray background.}
\end{figure*}
These experimentally obtained scalings of the loss rate and emittance growth with the excitation amplitude can be then used for the actual specification of the hollow electron lens.

@@

		reference bunch,\\ no excitation

@@

		excited bunch,\\ $10^{\mathrm{th}}$~turn V		\includegraphics[width=1.0\linewidth]{profile_v_10thv_slot_1300.png}

@@

The described change in tune entailed a change in the driving resonances, from previously the $7Q_x$ resonances in 2016 to mainly the $7Q_x$ and $7Q_y$ resonance in 2017. This can be seen from the FMA analysis for the 2016 optics (see Fig.~\ref{fig:patternfma} bottom left and bottom right), which shows a strong increase in diffusion for $7^{\mathrm{th}}$~turn pulsing in H around the $7Q_x$ resonances and only small changes for pulsing in V. In case of the 2017 optics and tune, the $7Q_x$ as well as the $7Q_y$ resonances cross the tune footprint (see Fig.~\ref{fig:7th2017fma} top left) and are also excited by the $7^{\mathrm{th}}$~turn pulsing visible as an increase of the diffusion in tune around the corresponding resonance lines for pulsing only in H or only in V and thus also for H+V (see Fig.~\ref{fig:7th2017fma}). An increase in the diffusion in tune however much weaker than for the $7Q_x$ and $7Q_y$ resonances is also observed around the $7Q_x + 7Q_y$ resonances for pulsing in V and H+V suggesting also an excitation of this $14^{\mathrm{th}}$ order resonance in addition. Why it is not visible for pulsing only in H is not understood.
\begin{figure}[h]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		no excitation
		\includegraphics[width=1.0\linewidth]{2017injnocolc15o+19_6noerru_dp0_ord14_annotate_7th.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$7^{\mathrm{th}}$ H+V
		\includegraphics[width=1.0\linewidth]{2017injnocolc15o+19_6noerrut7skhv_dp0_ord7.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$7^{\mathrm{th}}$ H
		\includegraphics[width=1.0\linewidth]{2017injnocolc15o+19_6noerrut7skh_dp0_ord7.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$7^{\mathrm{th}}$ V
		\includegraphics[width=1.0\linewidth]{2017injnocolc15o+19_6noerrut7skv_dp0_ord7.png}
	\end{minipage}	
	\caption{\label{fig:7th2017fma} FMA analysis in frequency space without excitation (top left) and for $7^{\mathrm{th}}$~turn pulsing in H+V (top right), only in H (bottom left) and only in V (bottom right) based on the 2017 injection optics with no machine errors and a tune of (62.27,60.295). The excitation is 96~nrad in both planes. The $7Q_x$ and $7Q_y$ resonance are both excited and in addition to the 14th order $7Q_x+7Q_y$.}
\end{figure}

@@

Qualitatively, this behavior could be also confirmed experimentally during the 2017 experiments. Here excitation amplitude dependent emittance growth was only recorded in the plane of excitation with all excitation amplitude dependent emittance growth rates in a comparable range. As example the relative horizontal and vertical emittance for pulsing in V and in H+V are depicted in Fig.~\ref{fig:ranexp}. Note that as expected from simulations and despite the large excitation amplitude dependent emittance growth, the growth rates stays constant. There is explicitly no fast adjustment phase succeeded by a new equilibrium state as observed for the $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing.
\begin{figure}[h]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		reference bunch,\\ no excitation
		\includegraphics[width=1.0\linewidth]{profile_v_ranv_slot_1698.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		excited bunch,\\ random excitation V
		\includegraphics[width=1.0\linewidth]{profile_v_ranv_slot_1532.png}
	\end{minipage}
	\caption{\label{fig:ranexpprof}Vertical beam profiles measured with the Beam Synchrotron Radiation Telescope (BSRT) during the 2017 experiments. The profiles are taken at the end of the random excitation in V. For both bunches the transverse damper is not active. The distribution changes in the reference bunch (left), here slot 1698, are much less than for the bunch experiencing the maximum excitation (right), here slot 2696.}
\end{figure}
The changes in emittance are also directly visible in the beam profiles measured with the BSRT. As an example Fig.~\ref{fig:ranexpprof} shows the profile for a random excitation only in V in comparison to a reference bunch. The distribution also stays Gaussian in case of a random excitation in contrast to the $10^{\mathrm{th}}$~turn pulsing in which case it clearly assumes a non-Gaussian shape (see Fig.~\ref{fig:10thexpprof}).

>>>>

\section{Sources of residual field in the proton beam core region and first order estimates\label{sec:core}}
With the current HEL layout as shown in Fig.~\ref{fig:hel_layout}, parasitic kicks on the proton beam core can arise in the central region due to profile imperfections in the electron beam and at the entrance and exit of the HEL, where electron and proton beam overlap. Currently, no HEL is installed in the LHC, and the kick on the proton beam core must therefore be emulated with the help of other devices already installed in the LHC. For this purpose the kick is approximated to first order by a dipole kick. During the experiments presented in this paper, the dipole kick with the corresponding excitation pattern could then be applied with the LHC transverse damper as will be described in detail in Section~\ref{sec:adt}. As will be shown in Sec.~\ref{core:sec:1} and \ref{core:sec:2}, the expected kick to first order from the HEL bends is estimated to be~(Eq.~\ref{eqn:kick_bends}):
\begin{equation}
	\Delta x'_{\rm{bends}},\Delta y'_{\rm{bends}}\approx \SI{0.5}{nrad},
\end{equation}
and for the central region due to profile imperfections~(Eq.~\ref{eqn:kick_central}):
\begin{equation}
	\Delta x'_{\rm{central \ region}}, \Delta y'_{\rm{central \ region}} \approx \SI{20}{nrad} % this should be rounded to the same number of digits as the 0.5nrad
\end{equation}
where the current HEL design parameters are used (see Table~\ref{tab:hllhc_param}--\ref{tab:hel_param}). The contribution from the central region is clearly dominating.

@@

Using these measurements and calculations of the relative electric field strengths in the transverse plane (Fig.~\ref{core:fig:1}), we obtain an average field in the hole of
\begin{equation}
	E_{\mathrm{rel}}= 5\%
\end{equation}
For a maximum kick of $392$~nrad as derived in Eq.~\ref{eqn:helkick}, the estimated dipole kick amplitude at the proton beam core then is:
\begin{equation}\label{eqn:kick_central}
\Delta x'_{\rm{central \ region}}, \Delta y'_{\rm{central \ region}} = \SI{20}{nrad}
\end{equation}

@@

The ADT kicker deflection angle is calculated from the kicker geometry and the excitation voltage on the deflection plates. The excitation voltage used in the experiments is a complicated waveform, generated by a very complex hardware (digital signal processor, chain of transmission lines, low and high power amplifiers). The peak excitation voltage (and subsequently the kick) seen by the beam can be only indirectly measured using the probes on each kicker. The measured values are therefore believed to be known with an error margin of 10-15\% for the experiments in 2017 and 50\% in 2016. The estimate for 2017 is based on only a single measurement and in the future and with more time available the determination of the error would definitely have to be improved. The error in 2016 is estimated from the transverse damper operational parameters. Due to lack of the dedicated machine time no precision calibration was performed. The maximum kick strength reachable with the ADT at injection without risking saturation is 96~nrad. As an example Fig.~\ref{fig:fill_meas} shows the oscillation amplitude of each bunch in both H and V planes captured by the LHC real time transverse activity monitor during the 2017 experiments and an excitation with $8^{\mathrm{th}}$ turn pulsing. The measured activity is well in line with the expected excitation pattern.
\begin{figure*}[h]
	\centering
	\includegraphics[width=1.0\linewidth]{bunchfilling_measured.png}	
	\caption{\label{fig:fill_meas} Bunch by bunch excitation detected by the real time transverse activity monitor during the 2017 experiment and an excitation with 8th turn pulsing.}
\end{figure*}

@@

The sensitivity to $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing is also observed in absence of machine errors in which case the only source of non-linearities are sextupoles and octupoles~\cite{md_sim_hel_res_ex_fitterer} suggesting that latter are responsible for the observed sensitivity. The driven resonances are revealed by the FMA analysis shown in Fig.~\ref{fig:patternfma}. The $7Q_x$ resonance is excited in case of the $7^{\mathrm{th}}$~turn pulsing and the $10Q_x$ and $10Q_y$ resonance in case of the $10^{\mathrm{th}}$~turn pulsing. As octupoles can only drive even resonances, the source of the $7Q_x$ resonances are the sextupoles while the octupoles only generate the tune footprint. The other pulsing patterns do not exhibit any increase in losses or emittance growth without magnetic errors and their effect can be thus attributed to the magnetic field errors, implying also the seed chosen for the simulation.
\begin{figure*}[h]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		no excitation
		\includegraphics[width=1.0\linewidth]{2016injnocolc15o+19_6noerru_dp0_ord10_annotate.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$10^{\mathrm{th}}$ H+V
		\includegraphics[width=1.0\linewidth]{2016injnocolc15o+19_6noerrut10skhv_dp0_ord10.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$7^{\mathrm{th}}$ H
		\includegraphics[width=1.0\linewidth]{2016injnocolc15o+19_6noerrut7skh_dp0_ord7.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$7^{\mathrm{th}}$ V
		\includegraphics[width=1.0\linewidth]{2016injnocolc15o+19_6noerrut7skv_dp0_ord7.png}
	\end{minipage}
	\caption{\label{fig:patternfma} FMA analysis without excitation (top left) for $10^{\mathrm{th}}$~turn pulsing in H+V (top left) and for $7^{\mathrm{th}}$~pulsing only in H (bottom left) and only in V (bottom right) and  based on the 2016 injection optics with no machine errors and a tune of (64.28,59.31). The excitation is 120~nrad in the corresponding plane. The absence of a strong excitation of any resonance in case of $7^{\mathrm{th}}$~turn pulsing only in V and the strong excitation in case of pulsing only in H confirms the excitation of the $7Q_x$ resonance. For $10^{\mathrm{th}}$~turn pulsing there is in contrast no significant difference between pulsing only in H, only in V or in H+V (see Appendix~\ref{app:sec:10}, Fig.~\ref{app:fig:fma:10}).}
\end{figure*}

One may have noticed that the emittance in case of the $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing starts at an increased initial value and then stays almost constant during the entire simulation time, which usually points to an error in the chosen simulation setup. This behavior is however typical for any resonant excitation and can be ascribed to the adjustment of the beam distribution during the first $10^4$ turns to a new equilibrium. As the beam distribution is only dumped every $10^4$ turns, this initial fast adjustment manifests itself in an increased initial emittance value in Fig.~\ref{fig:patternsim}, and also any other simulations over $10^6$ turns presented in this paper. Besides an increased emittance, the new distribution also becomes non-Gaussian. This is illustrated by means of the residual in respect to the Gaussian distribution for $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing in Fig.~\ref{fig:patternhist}. This initial fast change of the beam distribution is not an artifact of the simulation, but has been, to the knowledge of the authors, also for the first time experimentally observed during the 2016 and 2017 LHC experiments presented in this paper. The timescale of this fast adjustment is in experiments however much slower than predicted in simulations --- on the timescale of minutes compared to seconds. 
\begin{figure*}[t]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$7^{\mathrm{th}}$ H+V
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_t7skhv_3_5um_hist_x.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$10^{\mathrm{th}}$ H+V
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_t10skhv_3_5um_hist_y.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		random H+V
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_ranhv_3_5um_hist_x.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		no excitation
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_3_5um_hist_y.png}
	\end{minipage}	
	\caption{\label{fig:patternhist} Horizontal beam distribution for $7^{\mathrm{th}}$~turn pulsing (top left), vertical beam distribution for $10^{\mathrm{th}}$~turn pulsing (top right) and horizontal (bottom left) and vertical (bottom right) beam distribution for random excitation from distribution tracking simulations based on the 2016 injection optics with $(Q_x,Q_y)=(64.28,59.31)$ and standard errors. The excitation is applied in both planes with an amplitude of 96~nrad. The residual is defined as the final distribution (here after $10^4$~turns) minus the initial distribution (here after $10^2$~turns) as \%, and the ratio as the final distribution divided by the initial distribution. The red line in the plot of the residual is the difference between the Gaussian fit of the distribution and the distribution itself.}
\end{figure*}

@@

\subsection{$10^{\mathrm{th}}$ turn pulsing\label{sec:simex10}}
The $10^{\mathrm{th}}$~turn pulsing pattern was tested in 2016 with an excitation in the vertical plane only. As this was the first time a resonant excitation was tested in the LHC, only in total 48~bunches could be injected in order to still guarantee the protection of the machine. To test different excitation amplitudes during one fill and have enough statistics, the excitation was therefore only applied in one plane. As described earlier, the experiment started with a period of 12~minutes without excitation in order to let the beam distribution fully adjust to its equilibrium state after injection. Afterwards the excitation was applied with the excitation scheme shown in Fig.~\ref{fig:fill} and a maximum excitation amplitude of $A_{\mathrm{max}}=5\Delta A = 48~$nrad for a total of 11~minutes. The excitation was then further increased to the maximum value reachable without saturation of $A_{\mathrm{max}}=5\Delta A = 96~$nrad and kept for 11~minutes. In both cases, the excitation induced the following changes:
\begin{enumerate}
	\item Increase of the loss rate with the excitation amplitude (Figs.~\ref{fig:10thexp}.
	\item Increase of the emittance growth with the excitation amplitude in the vertical plane but not in the horizontal plane (Fig.~\ref{fig:10thexp}).
	\item Change of the beam distribution for the excited bunches (Fig.~\ref{fig:10thexpprof}).	
\end{enumerate}
As loss rates, emittance growth and beam distribution stayed unchanged for the $4$~reference bunches, the above observations can be also truly associated with the effect of the $10^{\mathrm{th}}$~turn pulsing.
\begin{figure*}[h]
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[height=0.75\linewidth]{2016_bunch_intensity_v10th_no_damper_avg.png}
	\end{minipage}
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[height=0.75\linewidth]{2016_emith_avg_rel_v10th_no_damper.png}
	\end{minipage}	
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[height=0.75\linewidth]{2016_emitv_avg_rel_v10th_no_damper.png}
	\end{minipage}	
	\caption{\label{fig:10thexp} Relative intensity (top) and relative horizontal (bottom left) and vertical (bottom right) emittance averaged over the bunches experiencing the same excitation amplitude during the 2016 experiments and for $10^{\mathrm{th}}$~turn pulsing in~V. For all bunches the damper was not active. Relative refers here to the normalization to the initial value. For each maximum amplitude $A_{\mathrm{max}}=48$~nrad and 96~nrad indicated with black arrows, the excitation amplitude was linearly increased for each group of 4~bunches (see Fig.~\ref{fig:fill}). The different colored solid lines labeled with $n\Delta A$ are the average over the 4~bunches with the same excitation amplitude together with the $1\sigma$ standard deviation shown as envelope. The area with a blue background indicates the fast adjustment period of the beam distribution transitioning into a new equilibrium state highlighted with a gray background.}
\end{figure*}
These experimentally obtained scalings of the loss rate and emittance growth with the excitation amplitude can be then used for the actual specification of the hollow electron lens.

@@

		reference bunch, no excitation

@@

		excited bunch, $10^{\mathrm{th}}$~turn V		\includegraphics[width=1.0\linewidth]{profile_v_10thv_slot_1300.png}

@@

The described change in tune entailed a change in the driving resonances, from previously the $7Q_x$ resonances in 2016 to mainly the $7Q_x$ and $7Q_y$ resonance in 2017. This can be seen from the FMA analysis for the 2016 optics (see Fig.~\ref{fig:patternfma} bottom left and bottom right), which shows a strong increase in diffusion for $7^{\mathrm{th}}$~turn pulsing in H around the $7Q_x$ resonances and only small changes for pulsing in V. In case of the 2017 optics and tune, the $7Q_x$ as well as the $7Q_y$ resonances cross the tune footprint (see Fig.~\ref{fig:7th2017fma} top left) and are also excited by the $7^{\mathrm{th}}$~turn pulsing visible as an increase of the diffusion in tune around the corresponding resonance lines for pulsing only in H or only in V and thus also for H+V (see Fig.~\ref{fig:7th2017fma}). An increase in the diffusion in tune however much weaker than for the $7Q_x$ and $7Q_y$ resonances is also observed around the $7Q_x + 7Q_y$ resonances for pulsing in V and H+V suggesting also an excitation of this $14^{\mathrm{th}}$ order resonance in addition. Why it is not visible for pulsing only in H is not understood.
\begin{figure*}[h]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		no excitation
		\includegraphics[width=1.0\linewidth]{2017injnocolc15o+19_6noerru_dp0_ord14_annotate_7th.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$7^{\mathrm{th}}$ H+V
		\includegraphics[width=1.0\linewidth]{2017injnocolc15o+19_6noerrut7skhv_dp0_ord7.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$7^{\mathrm{th}}$ H
		\includegraphics[width=1.0\linewidth]{2017injnocolc15o+19_6noerrut7skh_dp0_ord7.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$7^{\mathrm{th}}$ V
		\includegraphics[width=1.0\linewidth]{2017injnocolc15o+19_6noerrut7skv_dp0_ord7.png}
	\end{minipage}	
	\caption{\label{fig:7th2017fma} FMA analysis in frequency space without excitation (top left) and for $7^{\mathrm{th}}$~turn pulsing in H+V (top right), only in H (bottom left) and only in V (bottom right) based on the 2017 injection optics with no machine errors and a tune of (62.27,60.295). The excitation is 96~nrad in both planes. The $7Q_x$ and $7Q_y$ resonance are both excited and in addition to the 14th order $7Q_x+7Q_y$.}
\end{figure*}

@@

Qualitatively, this behavior could be also confirmed experimentally during the 2017 experiments. Here excitation amplitude dependent emittance growth was only recorded in the plane of excitation with all excitation amplitude dependent emittance growth rates in a comparable range. As example the relative horizontal and vertical emittance for pulsing in V and in H+V are depicted in Fig.~\ref{fig:ranexp}. Note that as expected from simulations and despite the large excitation amplitude dependent emittance growth, the growth rates stays constant. There is explicitly no fast adjustment phase succeeded by a new equilibrium state as observed for the $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing.
\begin{figure*}[h]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		reference bunch,\\ no excitation
		\includegraphics[width=1.0\linewidth]{profile_v_ranv_slot_1698.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		excited bunch,\\ random excitation V
		\includegraphics[width=1.0\linewidth]{profile_v_ranv_slot_1532.png}
	\end{minipage}
	\caption{\label{fig:ranexpprof}Vertical beam profiles measured with the Beam Synchrotron Radiation Telescope (BSRT) during the 2017 experiments. The profiles are taken at the end of the random excitation in V. For both bunches the transverse damper is not active. The distribution changes in the reference bunch (left), here slot 1698, are much less than for the bunch experiencing the maximum excitation (right), here slot 2696.}
\end{figure*}
The changes in emittance are also directly visible in the beam profiles measured with the BSRT. As an example Fig.~\ref{fig:ranexpprof} shows the profile for a random excitation only in V in comparison to a reference bunch. The distribution also stays Gaussian in case of a random excitation in contrast to the $10^{\mathrm{th}}$~turn pulsing in which case it clearly assumes a non-Gaussian shape (see Fig.~\ref{fig:10thexpprof}).

\begin{figure*}[h]
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2017_scale_amp_ranh_lblshort.png}
	\end{minipage}	
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2017_scale_amp_ranv_lblshort.png}
	\end{minipage}	
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2017_scale_amp_ranhv_lblshort.png}
	\end{minipage}	
	\caption{\label{fig:ranexplossplane} Comparison of scaling of loss rates with excitation amplitude for a random excitation only in H (top), only in V (center) and in H+V (bottom) measured during the 2017 experiment. The excitation amplitude error is 15\% indicated by the horizontal error bars. The loss rate is considerably reduced for the bunches with transverse damper active and is approximately additive with the excitation plane, meaning that double the loss rate is observed for an excitation in H+V compared to only in H or only in V.}
\end{figure*}
The experimental results for a maximum excitation amplitude of $A_{\mathrm{max}}=1$~nrad and $A_{\mathrm{max}}=6$~nrad also confirm the in general small losses predicted in simulations. The losses then increase more dramatically around the time when the maximum excitation amplitude is increased to $A_{\mathrm{max}}=12$~nrad (see Fig.~\ref{fig:ranexpprof}). This is most likely the point when, due to the large emittance growth, the physical aperture starts cutting into the more central region of the distribution. Furthermore, the dependence of the loss rate on the excitation plane is additive, meaning that double the loss rate is observed for an excitation in H+V compared to only in H or only V. The loss rate is also considerably reduced for the bunches with transverse damper active independent of the plane the excitation is applied in (see Fig.~\ref{fig:ranexplossplane}). In comparison to the $7^{\mathrm{th}}$~turn pulsing, the loss rate for a random excitation is smaller for a maximum excitation amplitude of $A_{\mathrm{max}}=6$~nrad, but larger for $A_{\mathrm{max}}=12$~nrad (see Fig.~\ref{fig:ranexploss}). This indicates an increase of the diffusion mostly in the beam tails for $7^{\mathrm{th}}$~turn pulsing, while for a random excitation mostly the core is affected.
\begin{figure}[h]
	\centering
	\includegraphics[width=1.0\linewidth]{2017_scale_amp_7th_ranhv_lblshort.png}
	\caption{\label{fig:ranexploss}  Comparison of scaling of loss rates with excitation amplitude for a random excitation and $7^{\mathrm{th}}$~turn pulsing in H+V measured during the 2017 experiment. The excitation amplitude error is 15\% indicated by the horizontal error bars.}
\end{figure}
